---
title: "Group analysis: Consonants"
author: "Thomas Tienkamp ^1^, Rob van Son ^2^^3^,  Bence Halpern ^2^^3^^4^ <br /> ^1^ Center for Language and Cognition Groningen, University of Groningen, Groningen, The Netherlands <br> ^2^ Amsterdam Center for Language and Communication, University of Amsterdam, Amsterdam, The Netherlands <br> ^3^ Netherlands Cancer Institute, Amsterdam, The Netherlands <br> ^4^ Multimedia Computing Group, Delft University of Technology, Delft, The Netherlands"
date: "Last compiled on: `r date()`"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    toc_float: TRUE
---

# 1. Description
This code shows the group analysis for both consonants and vowels belonging to the paper: Objective speech outcomes after surgical treatment for oral cancer: An acoustic analysis of a spontaneous speech corpus containing 32.850 tokens.

The code is organised in subsections. Its headers specify the content of the chunk.

## Table with the analysed consonants
| Place →<br>Manner↓ | Bilabial | Labio- dental | Dental | Alveolar | Post- Alveolar | Velar |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Plosive | /b/ /p/ |  |  | /t/ /d/ |  | /k/ /g/ |
| Fricative |  | /f/ /v/ | /θ/ /ð/ | /s/ /z/ | /ʃ/ /ʒ/ |  |
| Affricate |  |  |  | /tʃ/ /dʒ/ |  |  |


# 2. Preliminaries
## Width
```{r, echo=TRUE}
options(width = 180)
```

## Packages
```{r}
# Specify packages
packages <- c("tidyverse", "lme4", "lmerTest", "visreg", "car", "multcomp", 
              "emmeans", "boot", "gplots", "lattice", 
              "sjPlot", "sjmisc", "broom.mixed")

# Load packages or install if they're not installed yet
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# Show package version
for (package_name in packages) {
    print(paste(package_name, packageVersion(package_name)))
}
```

## Working directory
```{r}
# If working in the UWP
#mywd <- "X:/My Desktop/all-projects/nki_youtube-phoneme/"
#datawd <- paste0(mywd,"data/")

#If working from home
mywd <- "C:/Users/Thomas/Desktop/nki_youtube-phoneme/"
datawd <- paste0(mywd,"data/")

```

## Load data and helper functions
```{r}
dat <- read.csv(paste0(datawd, "Analysed.Phonemes.csv"), sep=',', na.strings = "--undefined--")

# Load additional plotting functions
source(paste0(mywd,"analysis/myFunctions2.R"))
```


# 3. Data exploration
## Check and change variable types
```{r}
#Check variable types
str(dat)
head(dat)

# Group, Time, Manner, Place, Voiced, Man_Plc need to be converted from character to factor
cols <- c("Speaker", "Group", "Time", "Manner", "Place", "Voiced", "Man_Plc")
dat[cols] <- lapply(dat[cols], as.factor)

# COGint, Tiltint, COGmean, and Skewmean need to be converted from character to numerical
cols2 <- c("COGint", "Tiltint", "COGmean", "Skewmean")
dat[cols2] <- lapply(dat[cols2], as.numeric)

#Check
str(dat) # Looks good
summary(dat)
```

## Get Frequency tables
We'll visualise how many instances there are of each phoneme per Group.
```{r}
frequency_table <- dat %>%
  group_by(Group, Phoneme, Place) %>%
  count()
frequency_table # Only 6 instances of /Z/ in the typical group and 13 in the pathoglogical group. Might be problematic.
```


## Filter plosives and fricatives
Here, we'll filter the phonemes in terms of relevant acoustic parameters (burst frequency, spectral mean, spectral skewness). In general, this means a split between fricatives and plosives. All three parameters are analysed in affricates, so they are included in both selections.
```{r}
fricatives <- dat %>%
  filter(Manner != "Plosive") %>% # rather than manner == "Fricative" to also include the affricates
  as.data.frame()
fricatives <- droplevels(fricatives) # Drop unused factorial levels
plosives <- dat %>%
  filter(Manner != "Fricative") %>% # rather than manner == "Plosive" to also include the affricates
  as.data.frame()
plosives <- droplevels(plosives) # Drop unused factorial levels
```


## Center numerical variables
All numerical variables are centered around the mean. This is done over the whole dataset if the parameter is applicable to all phonemes (duration & tilt) or to the subsection.
```{r}
# All data
dat$Tiltint.c <- scale(dat$Tiltint, scale = FALSE)
dat$Duration.c <- scale(dat$Duration, scale = FALSE)
dat$Duration.c <- dat$Duration.c * 1000 #Turn into miliseconds

# for plosives
plosives$COGint.c <- scale(plosives$COGint, scale = FALSE)

#For Fricatives
fricatives$COGmean.c <- scale(fricatives$COGmean, scale = FALSE)
fricatives$Skewmean.c <- scale(fricatives$Skewmean, scale = FALSE)

```

## Plot numerical variables
All numerical variables are plotted to visualise the distribution
```{r}
# All data
par(mfrow = c(2, 2))
boxplot(dat$Tiltint.c)
hist(dat$Tiltint.c)
boxplot(dat$Duration.c)
hist(dat$Duration.c) # Outliers on the right side which brings a positive skew

# Plosives
par(mfrow = c(1, 2))
boxplot(plosives$COGint.c)
hist(plosives$COGint.c) # Relatively strong positive skew

#Fricatives
par(mfrow = c(2, 2))
boxplot(fricatives$COGmean.c)
hist(fricatives$COGmean.c)
boxplot(fricatives$Skewmean.c)
hist(fricatives$Skewmean.c) # little positive skew
```

## Check distribution of numerical variables
All numerical variables are checked for normality. However, no changes are made yet. If the residuals of the models follow a normal distribution, then the dependent variable does not have to be normally distributed.
```{r}
par(mfrow = c(2, 3))
qqPlot(dat$Duration.c, main = "Duration") # Some outliers on the right side cause the tail
qqPlot(dat$Tiltint.c, main = "Tilt") # Small tail
qqPlot(plosives$COGint.c, main = "Burst Frequency") # S distribution
qqPlot(fricatives$COGmean.c, main = "Spectral mean") # S distribution
qqPlot(fricatives$Skewmean.c, main = "Spectral skewness") #Some tails
```

# 4. Set and change necessary contrasts
Orthogonal sum-to-zero contrasts are applied to binary factorial predictors

### Whole dataset
```{r}
# Set group contrast
contrast <- cbind (c(+1/2, -1/2))
colnames(contrast) <- c("-Typ+Path") #-0.5 = Typical, 0.5 = Pathological
contrasts(dat$Group) <- contrast
contrasts(dat$Group)

# Set voicing contrast
contrast <- cbind (c(-1/2, +1/2))
colnames(contrast) <- c("-FALSE+TRUE") #-0.5 = Voiceless, 0.5 = Voiced
contrasts(dat$Voiced) <- contrast
contrasts(dat$Voiced)
```

### Fricatives
```{r}
# Set group contrast
contrast <- cbind (c(+1/2, -1/2))
colnames(contrast) <- c("-Typ+Path") #-0.5 = Typical, 0.5 = Pathological
contrasts(fricatives$Group) <- contrast
contrasts(fricatives$Group)

# Set voicing contrast
contrast <- cbind (c(-1/2, +1/2))
colnames(contrast) <- c("-FALSE+TRUE") #-0.5 = Voiceless, 0.5 = Voiced
contrasts(fricatives$Voiced) <- contrast
contrasts(fricatives$Voiced)
```

### Plosives
```{r}
# Set group contrast
contrast <- cbind (c(+1/2, -1/2))
colnames(contrast) <- c("-Typ+Path") #-0.5 = Typical, 0.5 = Pathological
contrasts(plosives$Group) <- contrast
contrasts(plosives$Group)

# Set voicing contrast
contrast <- cbind (c(-1/2, +1/2))
colnames(contrast) <- c("-FALSE+TRUE") #-0.5 = Voiceless, 0.5 = Voiced
contrasts(plosives$Voiced) <- contrast
contrasts(plosives$Voiced)
```
# 5. Spectral Tilt
One compensatory mechanism for oral cancer speakers is to speak louder, which is reflected in the spectral tilt. We are therefore interested in the main effect of Group on spectral tilt while controlling for other variables.

## With grouped manner and Place
Hypothesis model:
```{r}
t.model <- lmerTest::lmer (Tiltint.c ~ Group + Man_Plc + Voiced + 
                     (1|Speaker),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
```

### Add interaction between group and Man_Plc
```{r}
t.model1 <- lmerTest::lmer (Tiltint.c ~ Group * Man_Plc + Voiced + 
                     (1|Speaker),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
anova(t.model, t.model1, refit = T) # Interaction improves the model

```

### Add interaction between Group and Voicing
```{r}
t.model2 <- lmerTest::lmer (Tiltint.c ~ Group * Man_Plc + Voiced + Group*Voiced +
                     (1|Speaker),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
anova(t.model1, t.model2, refit = T) # Interaction is necessary
```

### Add triple interaction
```{r}
t.model3 <- lmerTest::lmer (Tiltint.c ~ Group * Man_Plc * Voiced +
                     (1|Speaker),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
anova(t.model2, t.model3, refit = T) # Triple interaction is better
```

### Add an intercept for Time
```{r}
t.model3.1 <- lmerTest::lmer (Tiltint.c ~ Group * Man_Plc * Voiced +
                     (1|Speaker) + (1|Time),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
anova(t.model3, t.model3.1, refit = F) # Intercept is necessary
```

### Add slope for voiced by speaker
```{r}
t.model3.2 <- lmerTest::lmer (Tiltint.c ~ Group * Man_Plc * Voiced +
                     (1 + Voiced|Speaker) + (1|Time),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
anova(t.model3.1, t.model3.2, refit = F) # Slope is necessary

```

### Add slope for voiced by time
It could be the case that voicing problems differ per time frame as a result of radiation
```{r}
t.model3.3 <- lmerTest::lmer (Tiltint.c ~ Group * Man_Plc * Voiced +
                     (1 + Voiced|Speaker) + (1 + Voiced|Time),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
anova(t.model3.2, t.model3.3, refit = F) # Improves the model, model 3.3 is our final model
```

## Check assumptions
```{r}
# Model 3.3: Triple interaction
# Tiltint.c ~ Group * Man_Plc * Voiced
par(mfrow = c(2, 2))
acf(resid(t.model3.3)) # Good
car::vif(t.model3.3) # Very strong VIF for Man_Plc and its interactions -> is it purely driven by the interactions?
plot(resid(t.model3.3), fitted(t.model3.3)) # some tails but OK?
qqnorm(resid(t.model3.3)) # Small tails on both sides
qqline(resid(t.model3.3)) 

#Check whether the high VIF is driven by the interactions:
car::vif(t.model2) # All good
```

### Remove unused models
First remove the unused models:
```{r}
remove(t.model, t.model1, t.model2, t.model3, t.model3.1, t.model3.2)
```


# 6. Duration
This model explores whether the duration of consonants differs between pathological and typical speakers when controlling for place, manner and voicing.
```{r}
duration.model <- lmerTest::lmer (Duration.c ~ Group + Man_Plc + Voiced + 
                                    (1|Speaker), 
                            data = dat, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
```

### Add interaction between Group and Man_Plc
```{r}
duration.model2 <- lmerTest::lmer (Duration.c ~ Group * Man_Plc + Voiced + 
                                    (1|Speaker), 
                            data = dat, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(duration.model, duration.model2, refit = T) # Model 2 is preferred
```

### Add interaction between Group and Voicing
```{r}
duration.model3 <- lmerTest::lmer (Duration.c ~ Group * Man_Plc + Voiced * Group + 
                                    (1|Speaker), 
                            data = dat, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(duration.model2, duration.model3, refit = T) # Interaction is necessary
```

### Add triple interaction
```{r}
duration.model4 <- lmerTest::lmer (Duration.c ~ Group * Man_Plc * Voiced + 
                                    (1|Speaker), 
                            data = dat, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(duration.model3, duration.model4, refit = T) # Interaction is necessary
```
### Add intercept for time
```{r}
duration.model4.1 <- lmerTest::lmer (Duration.c ~ Group * Man_Plc * Voiced + 
                                    (1|Speaker) + (1|Time), 
                            data = dat, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(duration.model4, duration.model4.1, refit = F) # Intercept is necessary
```

### Add slope for voicing per speaker
```{r}
duration.model4.2 <- lmerTest::lmer (Duration.c ~ Group * Man_Plc * Voiced + 
                                    (1 + Voiced|Speaker) + (1|Time), 
                            data = dat, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(duration.model4.1, duration.model4.2, refit = F) # Slope is necessary
```

### Add slope for voicing per time frame
```{r}
duration.model4.3 <- lmerTest::lmer (Duration.c ~ Group * Man_Plc * Voiced + 
                                    (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = dat, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(duration.model4.2, duration.model4.3, refit = F) # Slope is necessary
```

## Assumptions of the final model
```{r}
par(mfrow = c(2, 2))
acf(resid(duration.model4.3)) # OK
car::vif(duration.model4.3) # Man_Plc and the interactions show VIF > 5
plot(resid(duration.model4.3), fitted(duration.model4.3)) # very bad
qqnorm(resid(duration.model4.3)) # Right tail
qqline(resid(duration.model4.3))

#Check whether this is due to the interactions
car::vif(duration.model) # VIF < 1.2 so it is fine
```

### Check outlier influence by trimming the residuals
```{r}
duration2 <- dat[abs(scale(resid(duration.model4.3))) < 2, ]

(1 - (nrow(duration2))/nrow(dat)) * 100  # 3.10% removed
duration.model4.3.V2 <- lmerTest::lmer (Duration.c ~ Group * Man_Plc * Voiced + 
                                    (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = duration2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
par(mfrow = c(2, 2))
acf(resid(duration.model4.3.V2)) # OK
car::vif(duration.model4.3.V2) # OK (as determined by the other comparison)
plot(resid(duration.model4.3.V2), fitted(duration.model4.3.V2)) #OK
qqnorm(resid(duration.model4.3.V2)) # Reasonably OK
qqline(resid(duration.model4.3.V2))
```

### Remove unused models
```{r}
rm(duration.model, duration.model2, duration.model3, duration.model4, duration.model4.1, duration.model4.2, duration.model4.3)
```

# 7. Burst frequency
This model explores whether the burst frequency of plosives differs per phone group when controlling for voicing.

## Hypothesis model
```{r}
bf.model <- lmerTest::lmer (COGint.c ~ Group * Man_Plc + Voiced + 
                              (1|Speaker) + (1|Time), 
                            data = plosives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
```

## Add interaction with Voiced
Perhaps the BF is differentially affected by Voicing per phone group per group
```{r}
bf.model1 <- lmerTest::lmer (COGint.c ~ Group * Man_Plc * Voiced + 
                              (1|Speaker) + (1|Time), 
                            data = plosives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
anova(bf.model, bf.model1, refit = T) # Interaction is necessary
```

## Add a random slope for Voiced per speaker
```{r}
bf.model1.1 <- lmerTest::lmer (COGint.c ~ Group * Man_Plc * Voiced + 
                              (1 + Voiced|Speaker) + (1|Time), 
                            data = plosives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
anova(bf.model1, bf.model1.1, refit = F) # Slope is necessary
```

## Slope for voiced per time
```{r}
bf.model1.2 <- lmerTest::lmer (COGint.c ~ Group * Man_Plc * Voiced + 
                              (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = plosives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000))) 
anova(bf.model1.1, bf.model1.2) # Slope is necessary --> bf.model1.2 is our final model
```
## Check assumptions
```{r}
par(mfrow = c(2, 2))
acf(resid(bf.model1.2)) # OK
car::vif(bf.model1.2) # Group and Voiced show some VIF, check whether this is due to interactions
car::vif(bf.model) # VIF is fine
plot(resid(bf.model1.2), fitted(bf.model1.2)) # some heteroscedasticity
qqnorm(resid(bf.model1.2)) # Right tail
qqline(resid(bf.model1.2)) 
```
## Assess influence of outliers on normality of residuals and heteroscedasticity
### Refit model with trimmed residuals
```{r}
plosives2 <- plosives[abs(scale(resid(bf.model1.2))) < 2, ]

(1 - (nrow(plosives2))/nrow(plosives)) * 100  # 5.7% removed
bf.model1.2.v2 <- lmerTest::lmer (COGint.c ~ Group * Man_Plc * Voiced + 
                              (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = plosives2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000))) 
summary(bf.model1.2)$coef
summary(bf.model1.2.v2)$coef
```

### Check assumptions
Problems of the residuals and heteroscedasticity are still not solved.
```{r}
par(mfrow = c(2, 2))
acf(resid(bf.model1.2.v2)) # OK
car::vif(bf.model1.2.v2) # Group and Voiced show some VIF
plot(resid(bf.model1.2.v2), fitted(bf.model1.2.v2)) # some heteroscedasticity
qqnorm(resid(bf.model1.2.v2)) # Right tail
qqline(resid(bf.model1.2.v2))
```
## Refit model with a square-root-transformation
```{r}
# Log-transform BF data
plosives$COGint.sqrt <- sqrt(plosives$COGint)
plosives$COGint.log <- log(plosives$COGint)

#Compare distributions
par(mfrow = c(2, 2))
qqPlot(plosives$COGint.sqrt, main = "SQRT")
qqPlot(plosives$COGint, main = "Original")
qqPlot(plosives$COGint.log, main = "LOG")

#Recentre
plosives$COGint.sqrt.c <- scale(plosives$COGint.sqrt, scale = FALSE)
plosives$COGint.log.c <- scale(plosives$COGint.log, scale = FALSE)

#Refit the model
sqrt.bf.model1 <- lmerTest::lmer (COGint.sqrt.c ~ Group * Man_Plc * Voiced + 
                              (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = plosives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
```

### Check assumptions
```{r}
par(mfrow = c(2, 2))
acf(resid(sqrt.bf.model1)) # OK
car::vif(sqrt.bf.model1) # OK
plot(resid(sqrt.bf.model1), fitted(sqrt.bf.model1)) # OK
qqnorm(resid(sqrt.bf.model1)) # OK
qqline(resid(sqrt.bf.model1))

```

As the residuals of the non-transformed burst-frequency models were not normally distributed and showed forms of heteroscedasticity, we will report on the sqrt transformed model.

### Remove other models
```{r}
rm(bf.model, bf.model1, bf.model1.1, bf.model1.2, bf.model1.2.v2)
```


# 8. Spectral mean
This model explores whether the spectral mean of fricatives differs per phone group when controlling for voicing.
## Hypothesis model
```{r}
cog.model <- lmerTest::lmer (COGmean.c ~ Group * Man_Plc + Voiced + 
                               (1|Speaker) + (1|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa",
                                        optCtrl = list(maxfun = 100000)))
```

## Add interaction with voicing
```{r}
cog.model1 <- lmerTest::lmer (COGmean.c ~ Group * Man_Plc * Voiced + 
                               (1|Speaker) + (1|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(cog.model, cog.model1, refit = T) # Interaction is necessary
```

## Add slope for voiced per speaker
```{r}
cog.model1.1 <- lmerTest::lmer (COGmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(cog.model1, cog.model1.1, refit = F) # Slope is necessary
```

## Add slope for voiced per time frame
```{r}
cog.model1.2 <- lmerTest::lmer (COGmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

anova(cog.model1.1, cog.model1.2, refit = F) # Slope is necessary and also our final model
```

## Check assumptions
```{r}
par(mfrow = c(2, 2))
acf(resid(cog.model1.2)) # OK
car::vif(cog.model1.2) # Man_Plc and interactions show high VIF
car::vif(cog.model) # High VIF is due to interactions, so it's fine
plot(resid(cog.model1.2), fitted(cog.model1.2)) # some heteroscedasticity
qqnorm(resid(cog.model1.2)) # Small tails on both sides
qqline(resid(cog.model1.2))
```

## Assess outlier influence on normality of residuals and heteroscedasticity
### Refit model
```{r}
fricatives2 <- fricatives[abs(scale(resid(cog.model1.2))) < 2, ]

(1 - (nrow(fricatives2))/nrow(fricatives)) * 100  # 5.2% removed
cog.model1.2.v2 <- lmerTest::lmer (COGmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = fricatives2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
```
### Check assumptions
```{r}
par(mfrow = c(2, 2))
acf(resid(cog.model1.2.v2)) # OK
plot(resid(cog.model1.2.v2), fitted(cog.model1.2.v2)) # still some heteroscedasticity but fine
qqnorm(resid(cog.model1.2.v2)) # Small tails on right side
qqline(resid(cog.model1.2.v2))
```
## Refit model with a square-root transformation
```{r}
#Transformations
fricatives$COGmean.sqrt <- sqrt(fricatives$COGmean)
fricatives$COGmean.log <- log(fricatives$COGmean)

#Compare distributions
par(mfrow = c(2, 2))
qqPlot(fricatives$COGmean.sqrt, main = "SQRT") # Slope of the tail seems steeper, so we'll try this.
qqPlot(fricatives$COGmean, main = "Original")
qqPlot(fricatives$COGmean.log, main = "LOG")

# Recentre
fricatives$COGmean.sqrt.c <- scale(fricatives$COGmean.sqrt, scale = FALSE)
fricatives$COGmean.log.c <- scale(fricatives$COGmean.log, scale = FALSE)

#Refit
sqrt.cog.model1.2 <- lmerTest::lmer (COGmean.sqrt.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
```

### Check assumptions and compare to original
```{r}
par(mfrow = c(2, 2))
acf(resid(sqrt.cog.model1.2)) # OK
plot(resid(sqrt.cog.model1.2), fitted(sqrt.cog.model1.2)) # OK
qqnorm(resid(sqrt.cog.model1.2)) # Tails on both sides now
qqline(resid(sqrt.cog.model1.2))

par(mfrow = c(2, 2))
acf(resid(cog.model1.2.v2)) # OK
plot(resid(cog.model1.2.v2), fitted(cog.model1.2.v2)) # OK
qqnorm(resid(cog.model1.2.v2)) # Small tail on right side
qqline(resid(cog.model1.2.v2))
```

As the transformed data resulted in tails on both sides and trimming the residuals did not help, we'll use the original model

### Remove other models
```{r}
rm(cog.model, cog.model1, cog.model1.1, cog.model1.2, sqrt.cog.model1.2)
```


# 9. Skewness
## Hypothesis model
This model explores whether the spectral skewness of fricatives differs per phone group when controlling for voicing.
```{r}
skew.model <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc + Voiced + 
                               (1|Speaker) + (1|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa",
                                        optCtrl = list(maxfun = 100000)))
```

## Add interaction with voicing
```{r}
skew.model1 <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc * Voiced + 
                               (1|Speaker) + (1|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa",
                                        optCtrl = list(maxfun = 100000)))

anova(skew.model, skew.model1, refit = T) # Interaction is necessary
```

## Add slope for voicing per speaker
```{r}
skew.model1.1 <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc * Voiced + 
                               (1 + Voiced|Speaker) + (1|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa",
                                        optCtrl = list(maxfun = 100000)))
anova(skew.model1, skew.model1.1, refit = F) # Slope is necessary
```


## Add slope for voicing per time frame
```{r}
skew.model1.2 <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc * Voiced + 
                               (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = fricatives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa",
                                        optCtrl = list(maxfun = 100000)))
anova(skew.model1.1, skew.model1.2, refit = F) # Necessary and final model
```

## Check assumptions
```{r}
par(mfrow = c(2, 2))
acf(resid(skew.model1.2)) # OK
car::vif(skew.model1.2) # Man_Plc and interactions show high VIF
car::vif(skew.model) # High VIF is due to interactions, so it's fine
plot(resid(skew.model1.2), fitted(skew.model1.2)) # heteroscedastic
qqnorm(resid(skew.model1.2)) # tails on both sides
qqline(resid(skew.model1.2))
```
## Check outlier influence
### Refit model
```{r}
skewness2 <- fricatives[abs(scale(resid(skew.model1.2))) < 2, ]

(1 - (nrow(skewness2))/nrow(fricatives)) * 100  # 5.6% removed
skew.model1.2.v2 <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = skewness2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))
```
### Check assumptions again
```{r}
par(mfrow = c(2, 2))
acf(resid(skew.model1.2.v2)) # OK
plot(resid(skew.model1.2.v2), fitted(skew.model1.2.v2)) # OK
qqnorm(resid(skew.model1.2.v2)) # OK!
qqline(resid(skew.model1.2.v2))
```


# 10. Summaries of final models
Here we'll plot all the final model summaries
```{r}
tidy.tilt <- tidy(t.model3.3, effects = "fixed", conf.int=TRUE) %>%
    mutate(p.adjust = p.adjust(p.value, method = "fdr"))
print(tidy.tilt, n = 40)

tidy.duration <- tidy(duration.model4.3.V2, effects = "fixed", conf.int=TRUE) %>%
    mutate(p.adjust = p.adjust(p.value, method = "fdr"))
print(tidy.duration, n = 40)
```

# 11. Manual Multiple comparisons

## 11.1 Burst frequency

First, check the used reference level for the final model: sqrt.bf.model1
```{r}
levels(plosives$Man_Plc) # Alveolar plosives is reference level
```

### 11.1.1 Check which rows of the summary need to be extracted:

- From the summary, row 2, 6, and 10 need to be extracted

```{r}
# summary(sqrt.bf.model1) # Uncomment to check
summary(sqrt.bf.model1)$coefficients[2, 0:5] #Group-Typ+Path 
summary(sqrt.bf.model1)$coefficients[6, 0:5] #Voiced-FALSE+TRUE
summary(sqrt.bf.model1)$coefficients[10, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE

```
### 11.1.2 Refit with Bi_Pl as reference
```{r}
plosives$Man_Plc <- relevel(plosives$Man_Plc, ref = "Bi_Pl")
bi_pl <- lmerTest::lmer (COGint.sqrt.c ~ Group * Man_Plc * Voiced + 
                              (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = plosives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

#Worked
# summary(bi_pl) # Uncomment to check
summary(bi_pl)$coefficients[2, 0:5] #Group-Typ+Path 
summary(bi_pl)$coefficients[6, 0:5] #Voiced-FALSE+TRUE
summary(bi_pl)$coefficients[10, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE

```

### 11.1.2 Refit with PA_Af as reference

```{r}
plosives$Man_Plc <- relevel(plosives$Man_Plc, ref = "PA_Af")
PA_Af <- lmerTest::lmer (COGint.sqrt.c ~ Group * Man_Plc * Voiced + 
                              (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = plosives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(PA_Af)$coefficients[2, 0:5] #Group-Typ+Path 
summary(PA_Af)$coefficients[6, 0:5] #Voiced-FALSE+TRUE
summary(PA_Af)$coefficients[10, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```

### 11.1.2 Refit with Ve_Pl as reference

```{r}
plosives$Man_Plc <- relevel(plosives$Man_Plc, ref = "Ve_Pl")
Ve_Pl <- lmerTest::lmer (COGint.sqrt.c ~ Group * Man_Plc * Voiced + 
                              (1 + Voiced|Speaker) + (1 + Voiced|Time), 
                            data = plosives, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(Ve_Pl)$coefficients[2, 0:5] #Group-Typ+Path 
summary(Ve_Pl)$coefficients[6, 0:5] #Voiced-FALSE+TRUE
summary(Ve_Pl)$coefficients[10, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```


## 11.2 Spectral mean

- Rows 2, 7, and 12 need to be extracted

```{r}
levels(fricatives2$Man_Plc)

#summary(cog.model1.2.v2) # Uncomment to check which rows need to be extracted


# Values for the reference level: Al_Fr
summary(cog.model1.2.v2)$coefficients[2, 0:5] #Group-Typ+Path 
summary(cog.model1.2.v2)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(cog.model1.2.v2)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE

```

### 11.2.1 Refit with De_Fr as reference
```{r}
fricatives2$Man_Plc <- relevel(fricatives2$Man_Plc, ref = "De_Fr")
levels(fricatives2$Man_Plc)

De_Fr<- lmerTest::lmer (COGmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = fricatives2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(De_Fr)$coefficients[2, 0:5] #Group-Typ+Path 
summary(De_Fr)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(De_Fr)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```

### 11.2.2 Refit with LD_Fr as reference
```{r}
fricatives2$Man_Plc <- relevel(fricatives2$Man_Plc, ref = "LD_Fr")
levels(fricatives2$Man_Plc)

LD_Fr<- lmerTest::lmer (COGmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = fricatives2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(LD_Fr)$coefficients[2, 0:5] #Group-Typ+Path 
summary(LD_Fr)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(LD_Fr)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```

### 11.2.3 Refit with PA_Af as reference
```{r}
fricatives2$Man_Plc <- relevel(fricatives2$Man_Plc, ref = "PA_Af")
levels(fricatives2$Man_Plc)

PA_Af.cog<- lmerTest::lmer (COGmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = fricatives2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(PA_Af.cog)$coefficients[2, 0:5] #Group-Typ+Path 
summary(PA_Af.cog)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(PA_Af.cog)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```

### 11.2.4 Refit with PA_Fr as reference
```{r}
fricatives2$Man_Plc <- relevel(fricatives2$Man_Plc, ref = "PA_Fr")
levels(fricatives2$Man_Plc)

PA_Fr <- lmerTest::lmer (COGmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = fricatives2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(PA_Fr)$coefficients[2, 0:5] #Group-Typ+Path 
summary(PA_Fr)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(PA_Fr)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```



## 11.3 Spectral skewness

Row 2, 7 and 12 need to be extracted
```{r}
levels(skewness2$Man_Plc) #Al_Fr is reference level
#summary(skew.model1.2.v2) #Uncomment to check

summary(skew.model1.2.v2)$coefficients[2, 0:5] #Group-Typ+Path 
summary(skew.model1.2.v2)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(skew.model1.2.v2)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```



### 11.3.1 Refit with De_Fr as reference
```{r}
skewness2$Man_Plc <- relevel(skewness2$Man_Plc, ref = "De_Fr")
levels(skewness2$Man_Plc)


skew_De_Fr <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = skewness2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(skew_De_Fr)$coefficients[2, 0:5] #Group-Typ+Path 
summary(skew_De_Fr)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(skew_De_Fr)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```

### 11.3.2 Refit with LD_Fr as reference
```{r}
skewness2$Man_Plc <- relevel(skewness2$Man_Plc, ref = "LD_Fr")
levels(skewness2$Man_Plc)


skew_LD_Fr <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = skewness2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(skew_LD_Fr)$coefficients[2, 0:5] #Group-Typ+Path 
summary(skew_LD_Fr)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(skew_LD_Fr)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```

### 11.3.3 Refit with PA_Af as reference
```{r}
skewness2$Man_Plc <- relevel(skewness2$Man_Plc, ref = "PA_Af")
levels(skewness2$Man_Plc)


skew_PA_Af <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = skewness2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(skew_PA_Af)$coefficients[2, 0:5] #Group-Typ+Path 
summary(skew_PA_Af)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(skew_PA_Af)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```


### 11.3.4 Refit with PA_Fr as reference
```{r}
skewness2$Man_Plc <- relevel(skewness2$Man_Plc, ref = "PA_Fr")
levels(skewness2$Man_Plc)


skew_PA_Fr <- lmerTest::lmer (Skewmean.c ~ Group * Man_Plc * Voiced + 
                               (1+ Voiced|Speaker) + (1 + Voiced|Time), 
                            data = skewness2, 
                            REML = T, 
                            lmerControl(optimizer = "bobyqa", 
                                        optCtrl = list(maxfun = 100000)))

summary(skew_PA_Fr)$coefficients[2, 0:5] #Group-Typ+Path 
summary(skew_PA_Fr)$coefficients[7, 0:5] #Voiced-FALSE+TRUE
summary(skew_PA_Fr)$coefficients[12, 0:5] #Group-Typ+Path:Voiced-FALSE+TRUE
```

