---
title: "Group analysis: Vowels"
author: "Thomas Tienkamp ^1^, Rob van Son ^2^^3^,  Bence Halpern ^2^^3^^4^ <br> <br> ^1^ Center for Language and Cognition Groningen, University of Groningen, Groningen, The Netherlands <br> ^2^ Amsterdam Center for Language and Communication, University of Amsterdam, Amsterdam, The Netherlands <br> ^3^ Netherlands Cancer Institute, Amsterdam, The Netherlands <br> ^4^ Multimedia Computing Group, Delft University of Technology, Delft, The Netherlands"
date: "Last compiled on: `r date()`"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    toc_float: TRUE
---

# 1. Description
This code shows the group analysis for the corner vowels belonging to the paper: Objective speech outcomes after surgical treatment for oral cancer: An acoustic analysis of a spontaneous speech corpus containing 32.850 tokens.

The code is organised in subsections. Its headers specify the content of the chunk.

## Table with the analysed phones
|      | Front | Mid | Back |   |   |   |
|------|-------|-----|------|:-:|:-:|:-:|
| High | /i/   |     | /u/  |   |   |   |
| Mid  |       |     |      |   |   |   |
| Low  |       |     | /a/  |   |   |   |


# 2. Preliminaries
## Width
```{r, echo=TRUE}
options(width = 180)
```

## Packages
```{r}
# Specify packages
packages <- c("tidyverse", "lme4", "lmerTest", "visreg", "car", "multcomp", 
              "emmeans", "boot", "gplots", "lattice", 
              "sjPlot", "sjmisc", "ggpubr")

# Load packages or install if they're not installed yet
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# Show package version
for (package_name in packages) {
    print(paste(package_name, packageVersion(package_name)))
}
```

## Working directory
```{r}
# If working in the UWP
#mywd <- "X:/My Desktop/all-projects/nki_youtube-phoneme/"
#datawd <- paste0(mywd,"data/")

#If working from home
mywd <- "C:/Users/Thomas/Desktop/nki_youtube-phoneme/"
datawd <- paste0(mywd,"data/")

```

## Load data and helper functions
```{r}
dat <- read.csv(paste0(datawd, "Analysed.Vowels.csv"), sep=',', na.strings = "--undefined--")

# Load additional plotting functions
source(paste0(mywd,"analysis/myFunctions2.R"))
```

# 3. Data exploration
## Check and change variable types
```{r}
#Check variable types
str(dat)
head(dat)

# Speaker, Group, Time, Phoneme, Height, Front need to be converted from character to factor
cols <- c("Speaker", "Group", "Time", "Phoneme", "Height", "Front")
dat[cols] <- lapply(dat[cols], as.factor)

# F1mid, F2mid, and F3mid need to be converted from character to numerical
cols2 <- c("F1mid", "F2mid", "F3mid")
dat[cols2] <- lapply(dat[cols2], as.numeric)

#Check
str(dat) # Looks good
summary(dat)
```
## Get Frequency tables
We'll visualise how many instances there are of each phoneme per Group.
```{r}
frequency_table <- dat %>%
  group_by(Group, Phoneme) %>%
  count()
frequency_table
```

## Center numerical variables
All numerical variables are centered around the mean.
```{r}
dat$F1mid.c <- scale(dat$F1mid, scale = FALSE)
dat$F2mid.c <- scale(dat$F2mid, scale = FALSE)
dat$F3mid.c <- scale(dat$F3mid, scale = FALSE)
```

## Plot numerical variables
All numerical variables are plotted to visualise the distribution
```{r}
par(mfrow = c(2,3))
hist(dat$F1mid.c, main = "F1")
hist(dat$F2mid.c, main = "F2")
hist(dat$F3mid.c, main = "F3")
qqPlot(dat$F1mid.c, main = "F1") # Potentially problematic
qqPlot(dat$F2mid.c, main = "F2")
qqPlot(dat$F3mid.c, main = "F3")

boxplot(dat$F1mid) # Are these outliers produced by one group?
f1check <- dat %>%
  filter(F1mid > 1000) %>%
  group_by(Group, Phoneme) %>%
  count()
f1check # Equally divided between groups if sample size is taken into account

f1.plot <- ggplot(dat, 
       mapping = aes(x = Phoneme, y = F1mid, fill = Group),
       na.rm = TRUE) +
    geom_jitter(aes(colour = Group),
                alpha = 0.3) +
  theme_minimal()

f2.plot <- ggplot(dat, 
       mapping = aes(x = Phoneme, y = F2mid, fill = Group),
       na.rm = TRUE) +
    geom_jitter(aes(colour = Group),
                alpha = 0.3) +
  theme_minimal()

ggarrange(f1.plot, f2.plot,
          ncol = 2,
          nrow = 1,
          common.legend = TRUE,
          widths = 10)
```

## Remove F1 outliers
```{r}
dat <- dat %>%
  dplyr::filter(Phoneme %in% c("i", "u") & F1mid <= 1000 | Phoneme == "a")

par(mfrow = c(1,2))
hist(dat$F1mid.c, main = "F1")
qqPlot(dat$F1mid.c, main = "F1") # Potentially problematic
```


## Log-transform
```{r}
dat$F1mid.log <- log(dat$F1mid)
dat$F2mid.log <- log(dat$F2mid)
dat$F1mid.sqrt <- sqrt(dat$F1mid)
dat$F2mid.sqrt <- sqrt(dat$F2mid)

# Check distribution again
par(mfrow = c(2,2))
qqPlot(dat$F1mid.log, main = "F1 LOG") # Looks good
qqPlot(dat$F2mid.log, main = "F2 LOG")
qqPlot(dat$F1mid.sqrt, main = "F1 SQRT")
qqPlot(dat$F2mid.sqrt, main = "F2 SQRT") 

# Recentre both log transformations
dat$F1mid.log.c <- scale(dat$F1mid.log, scale = FALSE)
dat$F2mid.log.c <- scale(dat$F2mid.log, scale = FALSE)

```


# 4. Set and change necessary contrasts
Orthogonal sum-to-zero contrasts are applied to binary factorial predictors
```{r}
# Set group contrast
contrast <- cbind (c(+1/2, -1/2))
colnames(contrast) <- c("-Typ+Path") #-0.5 = Typical, 0.5 = Pathological
contrasts(dat$Group) <- contrast
contrasts(dat$Group)
```


# 5. Model for F1
```{r}
f1.model <- lmerTest::lmer (F1mid.c ~ Group * Phoneme  +
                     (1|Speaker) + (1|Time),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
```

## Check assumptions
```{r}
par(mfrow = c(2, 2))
acf(resid(f1.model)) # OK
car::vif(f1.model)
plot(resid(f1.model), fitted(f1.model)) # Bad
qqnorm(resid(f1.model)) # Bad
qqline(resid(f1.model))
```

## Trim residuals and refit
```{r}
f1.2 <- dat[abs(scale(resid(f1.model))) < 2, ]

(1 - (nrow(f1.2))/nrow(dat)) * 100  # 4.74 removed
f1model.v2 <- lmerTest::lmer (F1mid.c ~ Group * Phoneme + 
                     (1|Speaker) + (1|Time),
                    data = f1.2,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
```
### Check assumptions
```{r}
par(mfrow = c(2, 2))
acf(resid(f1model.v2)) # OK
plot(resid(f1model.v2), fitted(f1model.v2)) # OK
qqnorm(resid(f1model.v2)) # OK
qqline(resid(f1model.v2))
```


# 6. Model for F2
```{r}
f2.model1 <- lmerTest::lmer (F2mid.c ~ Group * Phoneme + 
                     (1|Speaker) + (1|Time),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
```

## Check assumptions
```{r}
par(mfrow = c(2, 2))
car::vif(f2.model1) # OK
acf(resid(f2.model1)) # OK
plot(resid(f2.model1), fitted(f2.model1)) # fine: considering there are three phone groups
qqnorm(resid(f2.model1)) # some tails
qqline(resid(f2.model1))
```

## Scale residuals
```{r}
f2.2 <- dat[abs(scale(resid(f2.model1))) < 2, ]

(1 - (nrow(f2.2))/nrow(dat)) * 100  # 4.88 removed
f2model1.v2 <- lmerTest::lmer (F2mid.c ~ Group * Phoneme + 
                     (1|Speaker) + (1|Time),
                    data = f2.2,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
```

### Check assumptions
```{r}
par(mfrow = c(2, 2))
car::vif(f2model1.v2) # OK
acf(resid(f2model1.v2)) # OK
plot(resid(f2model1.v2), fitted(f2model1.v2)) # fine: considering there are three phone groups
qqnorm(resid(f2model1.v2)) # some tails, but smaller
qqline(resid(f2model1.v2))
```

# Try with log transformed data
```{r}
logf2.model1 <- lmerTest::lmer (F2mid.log.c ~ Group * Phoneme + 
                     (1|Speaker) + (1|Time),
                    data = dat,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
```

## Check assumptions
```{r}
par(mfrow = c(2, 2))
car::vif(logf2.model1) # OK
acf(resid(logf2.model1)) # OK
plot(resid(logf2.model1), fitted(logf2.model1)) # fine: considering there are three phone groups
qqnorm(resid(logf2.model1)) # Worse than trimmed residual model with no transformation
qqline(resid(logf2.model1))
```

## Scale residuals and refit
```{r}
log.f2.2 <- dat[abs(scale(resid(logf2.model1))) < 2, ]

(1 - (nrow(log.f2.2))/nrow(dat)) * 100  # 4.85 removed
log.f2model1.v2 <- lmerTest::lmer (F2mid.log.c ~ Group * Phoneme + 
                     (1|Speaker) + (1|Time),
                    data = log.f2.2,
                    REML = T,
                    lmerControl(optimizer = "bobyqa", 
                                optCtrl = list(maxfun = 100000)))
```
### Check assumptions
```{r}
par(mfrow = c(2, 2))
car::vif(log.f2model1.v2) # OK
acf(resid(log.f2model1.v2)) # OK
plot(resid(log.f2model1.v2), fitted(log.f2model1.v2)) # fine: considering there are three phone groups
qqnorm(resid(log.f2model1.v2)) # OK
qqline(resid(log.f2model1.v2))
```

# 7. Multiple comparisons between the final models
```{r}
f1.summary <- emmeans(f1model.v2,list(pairwise~Group|Phoneme), adjust = "fdr")
f2.summary <- emmeans(log.f2model1.v2,list(pairwise~Group|Phoneme), adjust = "fdr")

f1.summary
f2.summary
```
