---
title: "within-female01"
author: "Thomas Tienkamp ^1^, Rob van Son ^2^^3^,  Bence Halpern ^2^^3^^4^ <br /> ^1^ Center for Language and Cognition Groningen, University of Groningen, Groningen, The Netherlands <br> ^2^ Amsterdam Center for Language and Communication, University of Amsterdam, Amsterdam, The Netherlands <br> ^3^ Netherlands Cancer Institute, Amsterdam, The Netherlands <br> ^4^ Multimedia Computing Group, Delft University of Technology, Delft, The Netherlands"
date: "Last compiled on: `r date()`"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    toc_float: TRUE
---
# 1. Description
This code shows the within speaker analysis for female01 for both consonants and vowels belonging to the paper: Objective speech outcomes after surgical treatment for oral cancer: An acoustic analysis of a spontaneous speech corpus containing 32.850 tokens.

The code is organised in subsections. Its headers specify the content of the chunk.

## Table with the analysed consonants
| Place →<br>Manner↓ | Bilabial | Labio- dental | Dental | Alveolar | Post- Alveolar | Velar |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Plosive | /b/ /p/ |  |  | /t/ /d/ |  | /k/ /g/ |
| Fricative |  | /f/ /v/ | /θ/ /ð/ | /s/ /z/ | /ʃ/ /ʒ/ |  |
| Affricate |  |  |  | /tʃ/ /dʒ/ |  |  |

## Table with the analysed vowels
|      | Front | Mid | Back |   |   |   |
|------|-------|-----|------|:-:|:-:|:-:|
| High |/i/   |     | /u/  |   |   |   |
| Mid  |       |     |      |   |   |   |
| Low  |       |     | /a/  |   |   |   |

# 2. Preliminaries
## Width
```{r, echo=TRUE}
options(width = 180)
```

## Packages
```{r}
# Specify packages
packages <- c("tidyverse", "lme4", "lmerTest", "visreg", "car", "multcomp", 
              "emmeans", "boot", "gplots", "lattice", 
              "sjPlot", "sjmisc", "broom")

# Load packages or install if they're not installed yet
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# Show package version
for (package_name in packages) {
    print(paste(package_name, packageVersion(package_name)))
}
```

## Working directory
```{r}
# If working in the UWP
#mywd <- "X:/My Desktop/all-projects/nki_youtube-phoneme/"
#datawd <- paste0(mywd,"data/")

#If working from home
mywd <- "C:/Users/Thomas/Desktop/nki_youtube-phoneme/"
datawd <- paste0(mywd,"data/")

```

## Load data and helper functions
```{r}
c.dat <- read.csv(paste0(datawd, "Analysed.Phonemes.csv"), sep=',', na.strings = "--undefined--")
v.dat <- read.csv(paste0(datawd, "Analysed.Vowels.csv"), sep=',', na.strings = "--undefined--")

# Load additional plotting functions
source(paste0(mywd,"analysis/myFunctions2.R"))
```

## Filter out female01
```{r}
consonants <- c.dat %>%
  filter(Speaker == "Female01")
vowels <- v.dat %>%
  filter(Speaker == "Female01")
```

# 3. Data manipulation
## Change variable types
```{r}
str(consonants)

# Time, Manner, Place, Voiced, Man_Plc need to be converted from character to factor
cols1 <- c("Speaker", "Group", "Time", "Manner", "Place", "Voiced", "Man_Plc")
consonants[cols1] <- lapply(consonants[cols1], as.factor)

# COGint, Tiltint, COGmean, and Skewmean need to be converted from character to numerical
cols2 <- c("COGint", "Tiltint", "COGmean", "Skewmean")
consonants[cols2] <- lapply(consonants[cols2], as.numeric)

str(vowels)
# Time, Phoneme, Height, Front need to be converted from character to factor
cols3 <- c("Speaker", "Time", "Phoneme", "Height", "Front")
vowels[cols3] <- lapply(vowels[cols3], as.factor)

# F1mid, F2mid, and F3mid need to be converted from character to numerical
cols4 <- c("F1mid", "F2mid", "F3mid")
vowels[cols4] <- lapply(vowels[cols4], as.numeric)

str(vowels)
str(consonants)
```

## Set factorial contrasts
### Consonants
```{r}
# Set group contrast
contrast <- cbind (c(-1/2, +1/2))
colnames(contrast) <- c("-T1+T2") #-0.5 = Time 1, 0.5 = 2
contrasts(consonants$Time) <- contrast
contrasts(consonants$Time)

# Set voicing contrast
contrast <- cbind (c(-1/2, +1/2))
colnames(contrast) <- c("-FALSE+TRUE") #-0.5 = Voiceless, 0.5 = Voiced
contrasts(consonants$Voiced) <- contrast
contrasts(consonants$Voiced)
```

### Vowels
```{r}
# Set group contrast
contrast <- cbind (c(-1/2, +1/2))
colnames(contrast) <- c("-T1+T2") #-0.5 = Time 0, 0.5 = T1
contrasts(vowels$Time) <- contrast
contrasts(vowels$Time)
```

## Filter plosives and fricatives
Here, we'll filter the phonemes in terms of relevant acoustic parameters (burst frequency, spectral mean, spectral skewness). In general, this means a split between fricatives and plosives. All three parameters are analysed in affricates, so they are included in both selections.
```{r}
fricatives <- consonants %>%
  filter(Manner != "Plosive") %>% # rather than manner == "Fricative" to also include the affricates
  as.data.frame()
fricatives <- droplevels(fricatives) # Drop unused factorial levels
plosives <- consonants %>%
  filter(Manner != "Fricative") %>% # rather than manner == "Plosive" to also include the affricates
  as.data.frame()
plosives <- droplevels(plosives) # Drop unused factorial levels
```

## Center numerical variables
All numerical variables are centered around the mean. This is done over the whole dataset if the parameter is applicable to all phonemes (duration & tilt) or to the subsection.

### Consonants
```{r}
# All data
consonants$Tiltint.c <- scale(consonants$Tiltint, scale = FALSE)
consonants$Duration.c <- scale(consonants$Duration, scale = FALSE)
consonants$Duration.c <- consonants$Duration.c * 1000 #Turn into miliseconds

# for plosives
plosives$COGint.c <- scale(plosives$COGint, scale = FALSE)

#For Fricatives
fricatives$COGmean.c <- scale(fricatives$COGmean, scale = FALSE)
fricatives$Skewmean.c <- scale(fricatives$Skewmean, scale = FALSE)

```

### Vowels
```{r}
vowels$F1mid.c <- scale(vowels$F1mid, scale = FALSE)
vowels$F2mid.c <- scale(vowels$F2mid, scale = FALSE)
```

## Plot variables to get an idea of the distribution
```{r}
#All data
par(mfrow = c(2, 2))
boxplot(consonants$Tiltint.c)
hist(consonants$Tiltint.c)
boxplot(consonants$Duration.c)
hist(consonants$Duration.c)

# Plosives
par(mfrow = c(1, 2))
boxplot(plosives$COGint.c)
hist(plosives$COGint.c) # Relatively strong positive skew

#Fricatives
par(mfrow = c(2, 2))
boxplot(fricatives$COGmean.c)
hist(fricatives$COGmean.c)
boxplot(fricatives$Skewmean.c)
hist(fricatives$Skewmean.c) # little positive skew

#Vowels
par(mfrow = c(2,2))
hist(vowels$F1mid.c, main = "F1")
hist(vowels$F2mid.c, main = "F2")
qqPlot(vowels$F1mid.c, main = "F1")
qqPlot(vowels$F2mid.c, main = "F2")
```

# 4. Check contrasts
```{r}
contrasts(vowels$Time)
contrasts(consonants$Time)
contrasts(consonants$Voiced)
```

# 5. Consonant: tilt
## 5.1 Models
```{r}
t0 <- lm(Tiltint.c ~ Time + Man_Plc + Voiced, data = consonants)
t1 <- lm(Tiltint.c ~ Time * Man_Plc + Voiced, data = consonants)
t2 <- lm(Tiltint.c ~ Time + Man_Plc * Voiced, data = consonants)
t3 <- lm(Tiltint.c ~ Time * Man_Plc * Voiced, data = consonants)
AIC(t0) - AIC(t1) # not Better
AIC(t0) - AIC(t2) # Better
AIC(t2) - AIC(t3) #Not better
```

## 5.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(t2)) # Good
plot(resid(t2), fitted(t2)) #OK
qqnorm(resid(t2)) # OK
qqline(resid(t2)) 
```


# 6. Consonant: duration
## 6.1 Models
```{r}
d0 <- lm(Duration.c ~ Time + Man_Plc + Voiced, data = consonants)
d1 <- lm(Duration.c ~ Time * Man_Plc + Voiced, data = consonants)
AIC(d0) - AIC(d1) # Interaction is unnecessary
d2 <- lm(Duration.c ~ Time + Man_Plc * Voiced, data = consonants)
AIC(d1) - AIC(d2) # Interaction is unnecessary
d3 <- lm(Duration.c ~ Time * Man_Plc * Voiced, data = consonants)
AIC(d0) - AIC(d3) # Interaction is unnecessary

```

## 6.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(d0)) # Good
plot(resid(d0), fitted(d0)) #OK
qqnorm(resid(d0)) # Tail
qqline(resid(d0)) 

```

## 6.3 Refit with residuals
```{r fig.width=12, fig.height=7}
duration2 <- consonants[abs(scale(resid(d0))) < 2, ]

(1 - (nrow(duration2))/nrow(consonants)) * 100  # 4.4% removed
d0v2 <- lm(Duration.c ~ Time + Man_Plc + Voiced, data = duration2)

par(mfrow = c(1, 3))
acf(resid(d0v2)) # Good
plot(resid(d0v2), fitted(d0v2)) #OK
qqnorm(resid(d0v2)) # Better and OK-ish
qqline(resid(d0v2))
```

# 7. Plosives: burst frequency
## 7.1 Models
```{r}
b1 <- lm(COGint.c ~ Time * Man_Plc + Voiced, data = plosives)
b2 <- lm(COGint.c ~ Time * Man_Plc * Voiced, data = plosives)
AIC(b1) - AIC(b2) # Better
```

## 7.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(b2)) # Good
plot(resid(b2), fitted(b2)) #OK
qqnorm(resid(b2)) #Tails
qqline(resid(b2))
```

```{r}
plosives2 <- plosives[abs(scale(resid(b2))) < 2.2, ]

(1 - (nrow(plosives2))/nrow(plosives)) * 100  # 4.6% removed
b2v2 <- lm(COGint.c ~ Time * Man_Plc * Voiced, data = plosives2)

par(mfrow = c(1, 3))
acf(resid(b2v2)) # Good
plot(resid(b2v2), fitted(b2v2)) #OK-ish
qqnorm(resid(b2v2)) # Better and OK-ish
qqline(resid(b2v2))
```


# 8. Fricatives: spectral mean
## 8.1 Models
```{r}
sm1 <- lm(COGmean.c ~ Time * Man_Plc + Voiced, data = fricatives)
sm2 <- lm(COGmean.c ~ Time * Man_Plc * Voiced, data = fricatives)

AIC(sm1) - AIC(sm2) # Not - Better

```

## 8.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(sm1)) # Good
plot(resid(sm1), fitted(sm1)) #Not ok
qqnorm(resid(sm1)) #Some tails
qqline(resid(sm1))
```

## 8.3 Refit with trimmed residuals
```{r fig.width=12, fig.height=7}
cog2 <- fricatives[abs(scale(resid(sm1))) < 2.01, ]

(1 - (nrow(cog2))/nrow(fricatives)) * 100  # 4.77% removed
sm1v2 <- lm(COGmean.c ~ Time * Man_Plc + Voiced, data = cog2)

par(mfrow = c(1, 3))
acf(resid(sm1v2)) # Good
plot(resid(sm1v2), fitted(sm1v2)) #Better
qqnorm(resid(sm1v2)) # Better
qqline(resid(sm1v2))
```


# 9. fricatives: skewness
##  9.1 Models
```{r}
sk1 <- lm(Skewmean.c ~ Time * Man_Plc + Voiced, data = fricatives)
sk2 <- lm(Skewmean.c ~ Time * Man_Plc * Voiced, data = fricatives)

AIC(sk1) - AIC(sk2) # Better
```

##  9.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(sk2)) # Good
plot(resid(sk2), fitted(sk2)) #Not ok
qqnorm(resid(sk2)) #Heavy tail
qqline(resid(sk2))
```
## 9.3 Refit with trimmed residuals
```{r fig.width=12, fig.height=7}
skew2 <- fricatives[abs(scale(resid(sk2))) < 2.2, ]

(1 - (nrow(skew2))/nrow(fricatives)) * 100  # 5.9% removed
sk2v2 <- lm(Skewmean.c ~ Time * Man_Plc * Voiced, data = skew2)

par(mfrow = c(1, 3))
acf(resid(sk2v2)) # Good
plot(resid(sk2v2), fitted(sk2v2)) # OK
qqnorm(resid(sk2v2)) # Still not good.
qqline(resid(sk2v2))
```

## 9.4 Refit with transformed data
```{r}
fricatives$Skewmean.log <- log(fricatives$Skewmean)
fricatives$Skewmean.sqrt <- sqrt(fricatives$Skewmean)
fricatives$Skewmean.log.c <- scale(fricatives$Skewmean.log, scale = FALSE)
fricatives$Skewmean.sqrt.c <- scale(fricatives$Skewmean.sqrt, scale = FALSE)

par(mfrow = c(1, 3))
qqPlot(fricatives$Skewmean, main = "OG")
qqPlot(fricatives$Skewmean.log.c, main = "LOG")
qqPlot(fricatives$Skewmean.sqrt.c, main = "SQRT") # Seems good as it was a right tail
```

```{r}
sk.sqrt <- lm(Skewmean.sqrt.c ~ Time * Man_Plc * Voiced, data = fricatives)

par(mfrow = c(1, 3))
acf(resid(sk.sqrt)) # Good
plot(resid(sk.sqrt), fitted(sk.sqrt)) # OK
qqnorm(resid(sk.sqrt)) # Better
qqline(resid(sk.sqrt))
```

# 10. vowels: F1
##  10.1 Model
```{r}
f1 <- lm(F1mid.c ~ Time * Phoneme, data = vowels)
```

##  10.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(f1)) # Good
plot(resid(f1), fitted(f1)) # OK as there are two differences in height
qqnorm(resid(f1)) # Outliers
qqline(resid(f1))
```

## 10.3 Trimmed residuals

```{r}
firstf2 <- vowels[abs(scale(resid(f1))) < 1.65, ]

(1 - (nrow(firstf2))/nrow(vowels)) * 100  # 4.85% removed
f1v2 <- lm(F1mid.c ~ Time * Phoneme, data = firstf2)

par(mfrow = c(1, 3))
acf(resid(f1v2)) # Good
plot(resid(f1v2), fitted(f1v2)) # OK
qqnorm(resid(f1v2)) # Better
qqline(resid(f1v2))
```


# 11. vowels: F2
## 11.1 Models
```{r}
f2 <- lm(F2mid.c ~ Time * Phoneme, data = vowels)
```

## 11.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(f2)) # Good
plot(resid(f2), fitted(f2)) # OK as there are two differences in height
qqnorm(resid(f2)) # Some outliers
qqline(resid(f2))
```
## 11.3 Refit with trimmed residuals
```{r fig.width=12, fig.height=7}
secondf2 <- vowels[abs(scale(resid(f2))) < 2.5, ]

(1 - (nrow(secondf2))/nrow(vowels)) * 100  # 2.91% removed
f2v2 <- lm(F2mid.c ~ Time * Phoneme, data = secondf2)

par(mfrow = c(1, 3))
acf(resid(f2v2)) # Good
plot(resid(f2v2), fitted(f2v2)) # OK
qqnorm(resid(f2v2)) # Better
qqline(resid(f2v2))
```


# 12. Summaries of final models
```{r}
tidy.tilt <- tidy(t2, conf.int = T) %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr"))
tidy.duration <- tidy(d0v2, conf.int = T) %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr"))

print(tidy.tilt, n = 40)
options(width = 200)
print(tidy.duration, n = 40)

bf_sum <- emmeans(b2v2, list(pairwise~Time|Man_Plc|Voiced), adjust = "fdr")
cog_sum <- emmeans(sm1v2, list(pairwise~Time|Man_Plc), adjust = "fdr")
skew_sum <- emmeans(sk.sqrt, list(pairwise~Time|Man_Plc|Voiced), adjust = "fdr")
f1sum <- emmeans(f1v2, list(pairwise~Time|Phoneme), adjust = "fdr")
f2sum <- emmeans(f2v2, list(pairwise~Time|Phoneme), adjust = "fdr")

bf_sum
cog_sum
skew_sum
f1sum
f2sum
```

# 13. Visualise significant effects
```{r}
bfplot <- visreg(b2v2, "Time", 
       by = "Man_Plc",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(bfplot)

smplot <- visreg(sm1v2, "Time", 
       by = "Man_Plc",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(smplot)

skplot <- visreg(sk.sqrt, "Time", 
       by = "Man_Plc",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(skplot)
```

# 14. Visualise non-significant effects
```{r}
f1plot <- visreg(f1v2, "Time", 
       by = "Phoneme",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(f1plot)

f2plot <- visreg(f2v2, "Time", 
       by = "Phoneme",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(f2plot)

durplot <- visreg(d0v2, "Time",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(durplot)

tiltplot <- visreg(t2, "Time",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(tiltplot)
```

