---
title: "within-male01"
author: "Thomas Tienkamp ^1^, Rob van Son ^2^^3^,  Bence Halpern ^2^^3^^4^ <br> <br> ^1^ Center for Language and Cognition Groningen, University of Groningen, Groningen, The Netherlands <br> ^2^ Amsterdam Center for Language and Communication, University of Amsterdam, Amsterdam, The Netherlands <br> ^3^ Netherlands Cancer Institute, Amsterdam, The Netherlands <br> ^4^ Multimedia Computing Group, Delft University of Technology, Delft, The Netherlands"
date: "Last compiled on: `r date()`"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    toc_float: TRUE
---
# 1. Description
This code shows the within speaker analysis for female01 for both consonants and vowels belonging to the paper: Objective speech outcomes after surgical treatment for oral cancer: An acoustic analysis of a spontaneous speech corpus containing 32.850 tokens.

The code is organised in subsections. Its headers specify the content of the chunk.

## Table with the analysed consonants
| Place →<br>Manner↓ | Bilabial | Labio- dental | Dental | Alveolar | Post- Alveolar | Velar |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Plosive | /b/ /p/ |  |  | /t/ /d/ |  | /k/ /g/ |
| Fricative |  | /f/ /v/ | /θ/ /ð/ | /s/ /z/ | /ʃ/ /ʒ/ |  |
| Affricate |  |  |  | /tʃ/ /dʒ/ |  |  |

## Table with the analysed vowels
|      | Front | Mid | Back |   |   |   |
|------|-------|-----|------|:-:|:-:|:-:|
| High |/i/   |     | /u/  |   |   |   |
| Mid  |       |     |      |   |   |   |
| Low  |       |     | /a/  |   |   |   |

# 2. Preliminaries
## Width
```{r, echo=TRUE}
options(width = 180)
```

## Packages
```{r}
# Specify packages
packages <- c("tidyverse", "lme4", "lmerTest", "visreg", "car", "multcomp", 
              "emmeans", "boot", "gplots", "lattice", 
              "sjPlot", "sjmisc", "broom")

# Load packages or install if they're not installed yet
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# Show package version
for (package_name in packages) {
    print(paste(package_name, packageVersion(package_name)))
}
```

## Working directory
```{r}
# If working in the UWP
#mywd <- "X:/My Desktop/all-projects/nki_youtube-phoneme/"
#datawd <- paste0(mywd,"data/")

#If working from home
mywd <- "C:/Users/Thomas/Desktop/nki_youtube-phoneme/"
datawd <- paste0(mywd,"data/")

```

## Load data and helper functions
```{r}
c.dat <- read.csv(paste0(datawd, "Analysed.Phonemes.csv"), sep=',', na.strings = "--undefined--")
v.dat <- read.csv(paste0(datawd, "Analysed.Vowels.csv"), sep=',', na.strings = "--undefined--")

# Load additional plotting functions
source(paste0(mywd,"analysis/myFunctions2.R"))
```

## Filter out female01
```{r}
consonants <- c.dat %>%
  filter(Speaker == "Male01")
vowels <- v.dat %>%
  filter(Speaker == "Male01")
```

# 3. Data manipulation
## Change variable types
```{r}
str(consonants)
str(vowels)

# Recode the Time variable in three time stamps:
# T1 = 0-6 months, T2 = 6-12 months, T3 = 12 months
# Thus: T3&T4 --> T1; T5 --> T2; T8&T9 --> T3
consonants <- consonants %>%
  mutate(time_new = case_when((Time == "T3") ~ "T1",
                              (Time == "T4") ~ "T1",
                            (Time == "T5") ~ "T2",
                            (Time == "T8") ~ "T3",
                            (Time == "T9") ~ "T3"
                            )
        )

vowels <- vowels %>%
  mutate(time_new = case_when((Time == "T3") ~ "T1",
                              (Time == "T4") ~ "T1",
                            (Time == "T5") ~ "T2",
                            (Time == "T8") ~ "T3",
                            (Time == "T9") ~ "T3"
                            )
         )

# Time, Manner, Place, Voiced, Man_Plc need to be converted from character to factor
cols1 <- c("Speaker", "Group", "Time", "Manner", "Place", "Voiced", "Man_Plc", "time_new")
consonants[cols1] <- lapply(consonants[cols1], as.factor)

# COGint, Tiltint, COGmean, and Skewmean need to be converted from character to numerical
cols2 <- c("COGint", "Tiltint", "COGmean", "Skewmean")
consonants[cols2] <- lapply(consonants[cols2], as.numeric)

# Time, Phoneme, Height, Front need to be converted from character to factor
cols3 <- c("Speaker", "Time", "Phoneme", "Height", "Front", "time_new")
vowels[cols3] <- lapply(vowels[cols3], as.factor)

# F1mid, F2mid, and F3mid need to be converted from character to numerical
cols4 <- c("F1mid", "F2mid", "F3mid")
vowels[cols4] <- lapply(vowels[cols4], as.numeric)

# Check: Looks good
str(vowels)
str(consonants)
```

## Set factorial contrasts
### Consonants
```{r}
# Set group contrast
contrast <- cbind (c(-1/3, +2/3, -1/3), # T1 = -1/3, T2 = +2/3, T3 = -1/3
                   c(-0.5, 0, +0.5)) # T1 = -0.5, T2 = 0, T3 = +0.5
colnames (contrast) <- c("-T1-3+2", "-T1+T3") 
contrasts (consonants$time_new) <- contrast
contrasts(consonants$time_new)

# Set voicing contrast
contrast <- cbind (c(-1/2, +1/2))
colnames(contrast) <- c("-FALSE+TRUE") #-0.5 = Voiceless, 0.5 = Voiced
contrasts(consonants$Voiced) <- contrast
contrasts(consonants$Voiced)
```

### Vowels
```{r}
# Set group contrast
contrast <- cbind (c(-1/3, +2/3, -1/3), # T1 = -1/3, T2 = +2/3, T3 = -1/3
                   c(-0.5, 0, +0.5)) # T1 = -0.5, T2 = 0, T3 = +0.5
colnames (contrast) <- c("-T1-3+2", "-T1+T3") 
contrasts (vowels$time_new) <- contrast
contrasts(vowels$time_new)
```

## Filter plosives and fricatives
Here, we'll filter the phonemes in terms of relevant acoustic parameters (burst frequency, spectral mean, spectral skewness). In general, this means a split between fricatives and plosives. All three parameters are analysed in affricates, so they are included in both selections.
```{r}
fricatives <- consonants %>%
  filter(Manner != "Plosive") %>% # rather than manner == "Fricative" to also include the affricates
  as.data.frame()
fricatives <- droplevels(fricatives) # Drop unused factorial levels
plosives <- consonants %>%
  filter(Manner != "Fricative") %>% # rather than manner == "Plosive" to also include the affricates
  as.data.frame()
plosives <- droplevels(plosives) # Drop unused factorial levels
```

## Center numerical variables
All numerical variables are centered around the mean. This is done over the whole dataset if the parameter is applicable to all phonemes (duration & tilt) or to the subsection.

### Consonants
```{r}
# All data
consonants$Tiltint.c <- scale(consonants$Tiltint, scale = FALSE)
consonants$Duration.c <- scale(consonants$Duration, scale = FALSE)
consonants$Duration.c <- consonants$Duration.c * 1000 #Turn into miliseconds

# for plosives
plosives$COGint.c <- scale(plosives$COGint, scale = FALSE)

#For Fricatives
fricatives$COGmean.c <- scale(fricatives$COGmean, scale = FALSE)
fricatives$Skewmean.c <- scale(fricatives$Skewmean, scale = FALSE)

```

### Vowels
```{r}
#Filter unrealistic F1 values for /i/ and /u/
vowels <- vowels %>%
  dplyr::filter(Phoneme %in% c("i", "u") & F1mid <= 1000 | Phoneme == "a")

vowels$F1mid.c <- scale(vowels$F1mid, scale = FALSE)
vowels$F2mid.c <- scale(vowels$F2mid, scale = FALSE)
```

## Plot variables to get an idea of the distribution
```{r}
#All data
par(mfrow = c(2, 2))
boxplot(consonants$Tiltint.c)
hist(consonants$Tiltint.c)
boxplot(consonants$Duration.c)
hist(consonants$Duration.c)

# Plosives
par(mfrow = c(1, 2))
boxplot(plosives$COGint.c)
hist(plosives$COGint.c) # Relatively strong positive skew

#Fricatives
par(mfrow = c(2, 2))
boxplot(fricatives$COGmean.c)
hist(fricatives$COGmean.c) # Little skew
boxplot(fricatives$Skewmean.c)
hist(fricatives$Skewmean.c) # little positive skew

#Vowels
par(mfrow = c(2,2))
hist(vowels$F1mid.c, main = "F1")
hist(vowels$F2mid.c, main = "F2")
qqPlot(vowels$F1mid.c, main = "F1")
qqPlot(vowels$F2mid.c, main = "F2")
```

# 4. Check contrasts
```{r}
contrasts(vowels$time_new)
contrasts(consonants$time_new)
contrasts(consonants$Voiced)
```

# 5. Consonant: tilt
## 5.1 Models
```{r}
t0 <- lm(Tiltint.c ~ time_new + Man_Plc + Voiced, data = consonants)
t1 <- lm(Tiltint.c ~ time_new * Man_Plc + Voiced, data = consonants)
t2 <- lm(Tiltint.c ~ time_new * Man_Plc * Voiced, data = consonants)
AIC(t0) - AIC(t1) # Better
AIC(t1) - AIC(t2) # Better
```

## 5.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(t2)) # Good
plot(resid(t2), fitted(t2)) # OK
qqnorm(resid(t2)) # Small tails
qqline(resid(t2)) 
```

## 5.3 Refit with trimmed residuals
```{r fig.width=12, fig.height=7}
tilt2 <- consonants[abs(scale(resid(t2))) < 2, ]

(1 - (nrow(tilt2))/nrow(consonants)) * 100  # 4.93% removed
t2v2 <- lm(Tiltint.c ~ time_new * Man_Plc * Voiced, data = tilt2)

par(mfrow = c(1, 3))
acf(resid(t2v2)) # Good
plot(resid(t2v2), fitted(t2v2)) # OK
qqnorm(resid(t2v2)) # A bit better and reasonably OK
qqline(resid(t2v2))
```

# 6. Consonant: duration
## 6.1 Models
```{r}
d0 <- lm(Duration.c ~ time_new + Man_Plc + Voiced, data = consonants)
d1 <- lm(Duration.c ~ time_new * Man_Plc + Voiced, data = consonants)
d2 <- lm(Duration.c ~ time_new * Man_Plc * Voiced, data = consonants)
AIC(d0) - AIC(d1) # Better
AIC(d1) - AIC(d2) # Better
```

## 6.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(d2)) # Good
plot(resid(d2), fitted(d2)) #OK
qqnorm(resid(d2)) # Not ok
qqline(resid(d2)) 
```
## 6.3 Refit with trimmed residuals
```{r fig.width=12, fig.height=7}
duration2 <- consonants[abs(scale(resid(d2))) < 2, ]

(1 - (nrow(duration2))/nrow(consonants)) * 100  #4.5% removed
d2v2 <- lm(Duration.c ~ time_new + Man_Plc + Voiced, data = duration2)

par(mfrow = c(1, 3))
acf(resid(d2v2)) # Good
plot(resid(d2v2), fitted(d2v2)) # OK
qqnorm(resid(d2v2)) # Still some tails
qqline(resid(d2v2))
```

## 6.4 Refit with log transformation
```{r}
# Transform and recentre
consonants$Duration.log <- log(consonants$Duration)
consonants$Duration.sqrt <- sqrt(consonants$Duration)
consonants$Duration.log.c <- scale(consonants$Duration.log, scale = FALSE)
consonants$Duration.sqrt.c <- scale(consonants$Duration.sqrt, scale = FALSE)

par(mfrow = c(1, 3))
qqPlot(consonants$Duration, main = "OG")
qqPlot(consonants$Duration.log.c, main = "LOG")
qqPlot(consonants$Duration.sqrt.c, main = "SQRT") # Seems best
```

```{r fig.width=12, fig.height=7}
d3 <- lm(Duration.sqrt.c ~ time_new * Man_Plc * Voiced, data = consonants)

par(mfrow = c(1, 3))
acf(resid(d3)) # Good
plot(resid(d3), fitted(d3)) #OK
qqnorm(resid(d3)) # Much better
qqline(resid(d3)) 
```

## 6.5 Refit with trimmed residuals
```{r}
duration3 <- consonants[abs(scale(resid(d3))) < 3, ]
(1 - (nrow(duration3))/nrow(consonants)) * 100  #0.41% removed

d3v2 <- lm(Duration.sqrt.c ~ time_new * Man_Plc * Voiced, data = duration3)

par(mfrow = c(1, 3))
acf(resid(d3v2)) # Good
plot(resid(d3v2), fitted(d3v2)) # OK
qqnorm(resid(d3v2)) # OK
qqline(resid(d3v2))

```

# 7. Plosives: burst frequency
## 7.1 Models
```{r}
b1 <- lm(COGint.c ~ time_new * Man_Plc + Voiced, data = plosives)
b2 <- lm(COGint.c ~ time_new * Man_Plc * Voiced, data = plosives)
AIC(b1) - AIC(b2) # Better
```

## 7.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(b2)) # Good
plot(resid(b2), fitted(b2)) #Not ok
qqnorm(resid(b2)) #Not ok
qqline(resid(b2))
```
## 7.3 Refit with transformed data
```{r fig.width=12, fig.height=7}
plosives$COGint.log <- log(plosives$COGint)
plosives$COGint.sqrt <- sqrt(plosives$COGint)
plosives$COGint.log.c <- scale(plosives$COGint.log, scale = FALSE)
plosives$COGint.sqrt.c <- scale(plosives$COGint.sqrt, scale = FALSE)

par(mfrow = c(1, 3))
qqPlot(plosives$COGint, main = "OG")
qqPlot(plosives$COGint.log.c, main = "LOG")
qqPlot(plosives$COGint.sqrt.c, main = "SQRT") # Seems best
```

```{r}
b3 <- lm(COGint.sqrt.c ~ time_new * Man_Plc * Voiced, data = plosives)
par(mfrow = c(1, 3))
acf(resid(b3)) # Good
plot(resid(b3), fitted(b3)) #OK
qqnorm(resid(b3)) # Reasonably OK
qqline(resid(b3))

```
## 7.4 trimmed residuals
```{r}
burst2 <- plosives[abs(scale(resid(b3))) < 1.9, ]
(1 - (nrow(burst2))/nrow(plosives)) * 100  #4.7% removed

b3v2 <- lm(COGint.sqrt.c ~ time_new * Man_Plc + Voiced, data = burst2)

par(mfrow = c(1, 3))
acf(resid(b3v2)) # We already showed that this is due to the row ordering rather than the data.
plot(resid(b3v2), fitted(b3v2)) # OK
qqnorm(resid(b3v2)) # A bit better
qqline(resid(b3v2))
```


# 8. Fricatives: spectral mean
## 8.1 Models
```{r}
sm0 <- lm(COGmean.c ~ time_new * Man_Plc + Voiced, data = fricatives)
sm1 <- lm(COGmean.c ~ time_new * Man_Plc * Voiced, data = fricatives)

AIC(sm0) - AIC(sm1) # Better
```

## 8.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(sm1)) # Some ACF
plot(resid(sm1), fitted(sm1)) #Not ok
qqnorm(resid(sm1)) #Tails
qqline(resid(sm1))

#Combat ACF
set.seed(42)
sample <- sample(nrow(fricatives))
fricatives.shuffle <- fricatives[sample, ]

sm0.2 <- lm(COGmean.c ~ time_new * Man_Plc * Voiced, data = fricatives.shuffle)
par(mfrow = c(1, 3))
acf(resid(sm0.2)) # ACF = gone while effects stay identical
plot(resid(sm0.2), fitted(sm0.2)) #Not ok
qqnorm(resid(sm0.2)) #Some tails
qqline(resid(sm0.2))
```

## 8.3 Refit with trimmed residuals
```{r}
mean2 <- fricatives[abs(scale(resid(sm1))) < 2.1, ]
(1 - (nrow(mean2))/nrow(fricatives)) * 100  #4.47% removed

sm1v2 <- lm(COGmean.c ~ time_new * Man_Plc * Voiced, data = mean2)

par(mfrow = c(1, 3))
acf(resid(sm1v2)) # We already showed that this is due to the row ordering rather than the data.
plot(resid(sm1v2), fitted(sm1v2)) # OK
qqnorm(resid(sm1v2)) # Not oK
qqline(resid(sm1v2))
```

## 8.4 Refit with transformed data
```{r}
fricatives$COGmean.log <- log(fricatives$COGmean)
fricatives$COGmean.sqrt <- sqrt(fricatives$COGmean)
fricatives$COGmean.log.c <- scale(fricatives$COGmean.log, scale = FALSE)
fricatives$COGmean.sqrt.c <- scale(fricatives$COGmean.sqrt, scale = FALSE)

par(mfrow = c(1, 3))
qqPlot(fricatives$COGmean, main = "OG")
qqPlot(fricatives$COGmean.log.c, main = "LOG")
qqPlot(fricatives$COGmean.sqrt.c, main = "SQRT") # Seems best
```

```{r}
smsqrt <- lm(COGmean.sqrt.c ~ time_new * Man_Plc * Voiced, data = fricatives)

par(mfrow = c(1, 3))
acf(resid(smsqrt)) # We already showed that this is due to the row ordering rather than the data.
plot(resid(smsqrt), fitted(smsqrt)) # OK
qqnorm(resid(smsqrt)) # Not oK
qqline(resid(smsqrt))
```

## Refit with trimmed residuals
```{r}
mean3 <- fricatives[abs(scale(resid(smsqrt))) < 2.1, ]
(1 - (nrow(mean3))/nrow(fricatives)) * 100  #4.99% removed

smsqrtv2 <- lm(COGmean.sqrt.c ~ time_new * Man_Plc * Voiced, data = mean3)

par(mfrow = c(1, 3))
acf(resid(smsqrtv2)) # We already showed that this is due to the row ordering rather than the data.
plot(resid(smsqrtv2), fitted(smsqrtv2)) # OK
qqnorm(resid(smsqrtv2)) # Not oK
qqline(resid(smsqrtv2))
```

## 8.5 Opt for the simpler model instead:
```{r}
mean4 <- fricatives[abs(scale(resid(sm0))) < 2.4, ]
(1 - (nrow(mean4))/nrow(fricatives)) * 100  #1.8% removed

sm0v2 <- lm(COGmean.c ~ time_new * Man_Plc + Voiced, data = mean4)

par(mfrow = c(1, 3))
acf(resid(sm0v2)) # We already showed that this is due to the row ordering rather than the data.
plot(resid(sm0v2), fitted(sm0v2)) # OK
qqnorm(resid(sm0v2)) # OK
qqline(resid(sm0v2))
```



# 9. fricatives: skewness
##  9.1 Models
```{r}
sk1 <- lm(Skewmean.c ~ time_new * Man_Plc + Voiced, data = fricatives)
sk2 <- lm(Skewmean.c ~ time_new * Man_Plc * Voiced, data = fricatives)

AIC(sk1) - AIC(sk2) # Better
```

##  9.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(sk2)) # Good
plot(resid(sk2), fitted(sk2)) #Not ok
qqnorm(resid(sk2)) #Heavy tail
qqline(resid(sk2))
```
## 9.3 Refit with trimmed residuals
```{r fig.width=12, fig.height=7}
skew2 <- fricatives[abs(scale(resid(sk2))) < 2.1, ]

(1 - (nrow(skew2))/nrow(fricatives)) * 100  # 5.1% removed
sk2v2 <- lm(Skewmean.c ~ time_new * Man_Plc + Voiced, data = skew2)

par(mfrow = c(1, 3))
acf(resid(sk2v2)) # Good
plot(resid(sk2v2), fitted(sk2v2)) # OK
qqnorm(resid(sk2v2)) # Better
qqline(resid(sk2v2))
```

## 9.4 Refit with transformed data
```{r}
fricatives$Skewmean.log <- log(fricatives$Skewmean)
fricatives$Skewmean.sqrt <- sqrt(fricatives$Skewmean)
fricatives$Skewmean.log.c <- scale(fricatives$Skewmean.log, scale = FALSE)
fricatives$Skewmean.sqrt.c <- scale(fricatives$Skewmean.sqrt, scale = FALSE)

par(mfrow = c(1, 3))
qqPlot(fricatives$Skewmean, main = "OG")
qqPlot(fricatives$Skewmean.log.c, main = "LOG")
qqPlot(fricatives$Skewmean.sqrt.c, main = "SQRT") # Seems best
```

```{r}
sk3 <- lm(Skewmean.sqrt.c ~ time_new * Man_Plc * Voiced, data = fricatives)
par(mfrow = c(1, 3))
acf(resid(sk3)) # Good
plot(resid(sk3), fitted(sk3)) # OK
qqnorm(resid(sk3)) # OK!
qqline(resid(sk3))
```
## 9.5 Refit with trimmed residuals
```{r}
skew3 <- fricatives[abs(scale(resid(sk2))) < 2.2, ]

(1 - (nrow(skew3))/nrow(fricatives)) * 100  # 4.55% removed
sk3v2 <- lm(Skewmean.sqrt.c ~ time_new * Man_Plc * Voiced, data = skew3)

par(mfrow = c(1, 3))
acf(resid(sk3v2)) # Good
plot(resid(sk3v2), fitted(sk3v2)) # OK
qqnorm(resid(sk3v2)) # Better
qqline(resid(sk3v2))
```


# 10. vowels: F1
##  10.1 Model
```{r}
f1 <- lm(F1mid.c ~ time_new * Phoneme, data = vowels)
```

##  10.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(f1)) # Good
plot(resid(f1), fitted(f1)) # OK as there are two differences in height
qqnorm(resid(f1)) # OK
qqline(resid(f1))
```

#10.3 Refit with transformed data
```{r}
vowels$F1mid.log <- log(vowels$F1mid)
vowels$F1mid.sqrt <- sqrt(vowels$F1mid)
vowels$F1mid.log.c <- scale(vowels$F1mid.log, scale = FALSE)
vowels$F1mid.sqrt.c <- scale(vowels$F1mid.sqrt, scale = FALSE)

par(mfrow = c(1, 3))
qqPlot(vowels$F1mid, main = "OG")
qqPlot(vowels$F1mid.log.c, main = "LOG")# Seems best
qqPlot(vowels$F1mid.sqrt.c, main = "SQRT") 


```

```{r}
f1.log <- lm(F1mid.log.c ~ time_new * Phoneme, data = vowels)

par(mfrow = c(1, 3))
acf(resid(f1.log)) # Good
plot(resid(f1.log), fitted(f1.log)) # OK as there are two differences in height
qqnorm(resid(f1.log))
qqline(resid(f1.log))

```
## 10.4 Trimmed residuals
```{r}
logf1.2 <- vowels[abs(scale(resid(f1.log))) < 2.1, ]

(1 - (nrow(logf1.2))/nrow(vowels)) * 100  # 4.86% removed
f1.logv2 <- lm(F1mid.log.c ~ time_new * Phoneme, data = logf1.2)

par(mfrow = c(1, 3))
acf(resid(f1.logv2)) # Good
plot(resid(f1.logv2), fitted(f1.logv2)) # OK
qqnorm(resid(f1.logv2)) # OK!
qqline(resid(f1.logv2))
```


# 11. vowels: F2
## 11.1 Models
```{r}
f2 <- lm(F2mid.c ~ time_new * Phoneme, data = vowels)
```

## 11.2 Assumptions
```{r fig.width=12, fig.height=7}
par(mfrow = c(1, 3))
acf(resid(f2)) # Good
plot(resid(f2), fitted(f2)) # OK
qqnorm(resid(f2)) # Right tail
qqline(resid(f2))
```

## 11.3 Refit with trimmed residuals
```{r fig.width=12, fig.height=7}
secondf2 <- vowels[abs(scale(resid(f2))) < 2, ]

(1 - (nrow(secondf2))/nrow(vowels)) * 100  # 5.54% removed
f2v2 <- lm(F2mid.c ~ time_new * Phoneme, data = secondf2)

par(mfrow = c(1, 3))
acf(resid(f2v2)) # Good
plot(resid(f2v2), fitted(f2v2)) # OK
qqnorm(resid(f2v2)) # Better
qqline(resid(f2v2))
```


# 12. Summaries of final models
```{r}
duration_sum <- emmeans(d3v2, list(pairwise~time_new), adjust = "fdr")
tilt_sum <- emmeans(t2v2, list(pairwise~time_new), adjust = "fdr")
bf_sum <-emmeans(b3v2, list(pairwise~time_new|Man_Plc), adjust = "fdr") #SQRT transformed and trimmed
cog_sum <- emmeans(sm0v2, list(pairwise~time_new|Man_Plc), adjust = "fdr")
skew_sum <- emmeans(sk3v2, list(pairwise~time_new|Man_Plc|Voiced), adjust = "fdr") #SQRT transformed and trimmed
f1_sum <- emmeans(f1.logv2, list(pairwise~time_new|Phoneme), adjust = "fdr")
f2_sum <- emmeans(f2v2, list(pairwise~time_new|Phoneme), adjust = "fdr")

duration_sum
tilt_sum
bf_sum
cog_sum
skew_sum
f1_sum
f2_sum
```

# 13. Visualise significant effects
```{r}
bfplot <- visreg(b3v2, "time_new", 
       by = "Man_Plc",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(bfplot)

smplot <- visreg(sm0v2, "time_new", 
       by = "Man_Plc",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(smplot)

skplot <- visreg(sk3v2, "time_new", 
       by = "Man_Plc",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(skplot)
```

# 14. Visualise non-significant effects
```{r}
f1plot <- visreg(f1.logv2, "time_new", 
       by = "Phoneme",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(f1plot)

f2plot <- visreg(f2v2, "time_new", 
       by = "Phoneme",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(f2plot)

durplot <- visreg(d3v2, "time_new",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(durplot)

tiltplot <- visreg(t2v2, "time_new",
       overaly = TRUE,
       gg=TRUE) + 
  scale_fill_brewer(palette="Paired") + 
  theme_minimal()
print(tiltplot)
```